Learning with Misspecified Models: the case of overestimation

### 2023-10-15 Paper/Intro.tex
Separate file with only intro

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2.5cm]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

@@

\newcommand{\EE}[1]{\mathbb{E}\left[#1\right]}
\newcommand{\PP}[1]{\mathbb{P}\left(#1\right)}
\newcommand{\CP}[2]{\mathbb{P}\left(#1 \,| \, #2 \right)}
\newcommand{\CE}[2]{\mathbb{E}\left[#1\,|\,#2 \right]}
\usepackage[bottom]{footmisc}
\usepackage[doublespacing]{setspace}
\usepackage[normal]{caption}
\usepackage{dsfont}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{hyperref}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Learning with Mental Models: the case of overestimation},
  pdfauthor={Jimena Galindo},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Learning with Mental Models: the case of overestimation}
\author{Jimena Galindo}
\date{October 15, 2023}

@@

\hypertarget{theoretical-framework}{%
\section{Theoretical Framework}\label{theoretical-framework}}

The theories that I consider make predictions in two distinct
frameworks. I first detail each theory within their original framework
and then develop a unifying example that allows me to compare the
predictions of all of them. In framework 1 I consider a setting in which
the an agent observes a continuous output and infers the underlying
state. In framework 2, the agent observes a binary outcome and infers
the underlying probability of success. In both frameworks, I consider
two nested theories of learning.

\hypertarget{framework-1}{%
\subsection{Framework 1}\label{framework-1}}

An agent is of type \(\theta \in \Theta\) and faces an unknown exogenous
state \(\omega\) drawn from some density \(f\) over \(\Omega\). The
agent knows the distribution of \(\omega\) but not its realized value.
His belief about the state, \(p_0(\omega)\), coincide with the true
distribution, \(f\); and his belief about the type is \(p_0(\theta)\).
Let the agent's true type be \(\theta^{*}\), and the realized state be
\(\omega^{*}\).

An agent's belief about the type is \emph{misspecified} if it assigns
probability zero to their true type. Furthermore, the agent is
\emph{dogmatic} if he holds a degenerate belief that places probability
one on being of type \(\hat{\theta}\). An agent can be dogmatic and
misspecified, in which case \(\hat{\theta} \neq \theta^*\) and
\(p_0(\hat{\theta}) = 1\).

The agent chooses an action \(a\in A\) and observes a noisy outcome
\(h\). The outcome is a function of the agent's type, the state, and the
action. In particular \(h = h(\theta^*, \omega^*, a) + \varepsilon\)
with \(h(\cdot)\) increasing in both \(\theta^*\) and \(\omega^*\), and
such that conditional on a pair of parameters \((\theta, \omega)\),
there is a unique optimal action. additionally,
\(\varepsilon\sim N(0, \sigma)\) is noise in the output.

After observing the outcome, the agent updates his beliefs about
\(\theta\) and \(\omega\) using some algorithm and moves on to the next
period. He repeats this process infinitely many times. I make the
simplifying assumption that the agent is myopic and chooses the action
that maximizes the payoff in each period. This assumption simplifies the
analysis and plays a role in whether an agent who updates their beliefs
using Bayes rule would learn the truth or not, however, for the main
theory discussed in this section, the results hold even when relaxing
this assumption. The behavior of a forward-looking agent is further
discussed in the conclusion.

A key notion in this setting is that of a self-defeating
equilibrium\footnote{This notion is an adaptation of the Berk-Nash equilibrium in 
\citet{Esponda2016} to this setting with only one agent}. A
\emph{self-defeating equilibrium} is a belief and action pair such that
the agent's belief about their type is misspecified and the outcome
generated by the action is consistent with the misspecified belief. This
means that the average outcome under the true type and the true state
equals the average output the agent expects under the misspecified
belief. The agent's belief is said to be \emph{stable} when this
happens.

Within this framework, I consider two nested theories of belief
updating: the first one is a dogmatic modeler from \citet{Heidhues2018},
The second one is a switcher, as in \citet{Ba2023}. The dogmatic modeler
can be seen as a switcher with an infinitely sticky initial belief so
the two theories are nested. However, outside of the limiting case, they
produce different predictions about the agent's behavior. I discuss each
of them separately in what follows.

\hypertarget{the-dogmatic-modeler}{%
\subsubsection{The Dogmatic Modeler}\label{the-dogmatic-modeler}}

Take the action space to be the same as the state space \(\Omega\) and
\(\Omega =A=[\underline{a}, \infty)\).

A dogmatic agent does not update their beliefs about \(\theta\),
instead, he holds a degenerate belief that places probability on
\(\hat{\theta}\), which is potentially misspecified. In this case, no
matter how much information he gathers against being of type
\(\hat{\theta}\), he will not update his beliefs about it. Any
discrepancies between the observed outcomes and his believed type are
incorporated using the Bayes rule to update his beliefs about
\(\omega\). \citet{Heidhues2018} show that, under certain assumptions on
the per-period utility,\footnote{The assumptions are that 
$u$ is twice continuously differentiable with: (i)$u_{aa}<0$ and $u_a(\underline{a},  \theta, \omega)>0>u_a(\Bar{a}, \theta, \omega)$, 
(ii) $u_{\theta}, u_{\omega}>0$ and (iii) $u_{a\theta}<0$ and $u_{a\omega}>0$. Where $\underline{a}$ is the
minimal action and $\Bar{a}$ is the maximal one. 
 The direction of the derivatives is a normalization
and the results would hold even when the signs are reversed.} a dogmatic
modeler will inevitably fall into a self-defeating equilibrium. The
equilibrium will be such that the outcomes they observe reinforce their
belief on \(\omega\) in such a way that as \(t\to\infty\) the agent will
be sure that the state is some \(\omega^{'}\) consistent with their
believed type and the observed data. In other words, they will be in a
self-defeating equilibrium with a stable belief that places probability
one on the incorrect parameters \((\hat{\theta}, \omega_{\infty})\)

The mechanism by which the dogmatic agent falls into the self-defeating
equilibrium is the following: Suppose the agent holds the misspecified
belief that they are type \(\hat{\theta}>\theta^*\). For any prior over
\(\omega\), the agent will be disappointed by the outcome. He expected a
gain of \(h(\hat{\theta}, \mathbb{E}(\omega), a)\) but instead observes
\(h(\hat{\theta}, \omega^*, a)\). There are two possible sources for the
disappointment: the first is that the realized state is lower than the
expected state; the second source is that the agent is of type
\(\theta^*\) and therefore, for all possible states, his gain will be
lower than what he expected. Because the agent is dogmatic, he will not
update his beliefs about \(\theta\) and as a consequence will attribute
the disappointment to the state being lower than expected. He will
continue to update in this way until he converges to a belief about
\(\omega\) that is stable. Such a belief will explain the observed
utility perfectly and allow the agent to rationalize his dogmatic belief
about \(\theta\). Under the assumptions of \citet{Heidhues2018}, there
is a unique value of \(\omega\) at which the belief is stable, I will
refer to such value as \(\omega_\infty\). This mechanism is illustrated
in Example 1.

\textbf{Example 1: } \emph{Set \(A = \Omega=[\underline{a}, \infty)\)
and consider a student with intrinsic ability \(\theta^*\geq 0\) who
faces a grading procedure \(\omega^*\) that is unknown to them. The
student knows that a higher \(\omega^*\) is more likely to yield a
higher grade. In particular, she knows the grade is given by
\((\theta^*+a)\omega^*\).}

\emph{The student must choose an effort level \(a\), which determines
her grade. For whatever the chosen effort level, the student must pay a
cost \(c(a) = \frac{1}{2}a^2\) and observes a grade of
\((\theta^*+a)\omega^*\). She repeats this process for infinitely many
periods. Assume also that the student's prior is such that
\(\mathbb{E}[\omega]= \omega^*\) and she is dogmatic about being of type
\(\hat{\theta}>\theta^*\).\footnote{The example is illustrated for an overconfident student but 
the results are symmetric for a dogmatic student who initially 
places probability one on some $\tilde{\theta}<\theta^*$.} Therefore,
the student's payoff in period \(t\) is given by }

\begin{equation}
u_t(a_t; \theta^*, \omega^*) = (\theta^*+a_t)\omega^* - \frac{1}{2}a^2 + \varepsilon_t
\end{equation}

\emph{Under this specification, the myopic optimal effort level is
\(a_t^* = \omega^*\). Because the student does not know \(\omega^*\), he
will choose \(a_t = \mathbb{E}_t(\omega)\) where the expectation is
taken with respect to the student's belief at the beginning of period
\(t\). If she does not revise his effort choice for \(k\) periods, she
will receive an average utility of
\((\theta^*+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}\) but he was expecting
an average utility of
\((\hat{\theta}+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}\). In response, she
will apply Bayes rule to update his beliefs about \(\omega\) to get the
posterior belief with
\(\mathbb{E}_{t+k}[\omega] = \frac{(\theta^{*} + \omega^{*})\omega^{*}}{\hat{\theta} + \omega^{*}}\)
which is lower than the initial belief. This will cause the student to
choose a lower effort at \(t+k\). As a result, she will again receive an
average utility that is lower than what he expected which will cause her
belief to drift further down. This process will continue until the
average utility equals her expected utility under the dogmatic belief
that assigns probability 1 to \(\hat{\theta}\). At that point, the
student will have reached a self-defeating equilibrium and she will
continue to choose sub-optimal effort forever. }

Although the model of a dogmatic modeler rationalizes the prevalence of
overconfident (underconfident) beliefs, the assumption that the agent
has a degenerate belief and no mechanism through which he can update
such belief is very restrictive. An alternative approach is proposed by
\citet{Ba2023}. She proposes an extension of the dogmatic agent who is
able to swithch from one dogmatic belief to another. By doing so, the
agent can avoid the self-defeating equilibria and end up being dogmatic
and correctly specified.

\hypertarget{the-switcher}{%
\subsubsection{The Switcher}\label{the-switcher}}

An agent is a \emph{switcher} if they behave as a dogmatic, but are
willing to entertain the possibility that they are of a different type.
In particular, when they start off as a misspecified dogmatic, they are
willing to switch to a different dogmatic belief if the data is
convincing enough. Their prior is still degenerate and assigns
probability one to a particular type, and zero to all other types. This
means that a Bayesian update on \(\theta\) does not change their beliefs
about the type. However, they are willing to entertain two such beliefs
and have a mechanism by which they decide which belief to adopt at any
period \(t\).

In order to abandon their initial dogmatic belief, the agent needs to
observe a sequence of outcomes that are sufficiently unlikely to have
happened if they were of the type they initially believed. In order to
evaluate if the evidence is convincing enough, they keep track of the
likelihood that each of the possible types generated the data. If the
likelihood ratio is sufficiently large, the agent will switch to the
alternative and behave as if they are dogmatic about the new type.

In particular, for an agent that starts with a dogmatic belief that they
are of type \(\hat{\theta}\) but is willing to consider the alternative
explanation that they are of type \(\tilde{\theta}\), the agent will
switch to the alternative if:

\[\frac{p[h^t|\tilde{\theta}]}{p[h^t|\hat{\theta}]} > \alpha\geq 1\]

Where \(h^t\) is the history of outcomes up to time \(t\) and \(\alpha\)
is the switching
threshold.\footnote{Notice that if $\alpha \to \infty$, the behavior of the switcher will be indistinguishable from that of the Dogmatic modeler. 
In this sense, the switcher is a generalization of the dogmatic type.}
By keeping track of the likelihood ratio, the agent can perform a
\emph{Bayesian hypothesis test} and adopt the Dogmatic belief that best
fits the
data.\footnote{In a related problem \citet{Schwarstein2021} proposes a similar updating procedure which relies on the Bayesian 
hypothesis test. However, in their model there is a sender who optimally chooses to propose a model that fits the data}

By allowing the agent to keep track of the likelihoods and switching to
an alternative type, the switcher can avoid the self-confirming
equilibria. However, if the prior belief on \(omega\) is sufficiently
tight around a self-defeating equilibrium, the switcher might look
identical to the dogmatic even in a case where \(\alpha\) is not too
large. This happens because under the agent's prior, the likelihood
ratio is unlikely to grow as fast as it is needed to escape the
self-defeating equilibrium. In such situations, we say that the
misspecified belief is persistent.

\hypertarget{framework-2}{%
\subsection{Framework 2}\label{framework-2}}

As in framework 1, the agent is of some type \(\theta^* \in \Theta\) and
the state is \(\omega \sim F(\Omega)\). In this case, the agent chooses
an action \(a\in A\) and observes a binary outcome that is either a
success or a failure. Denote the outcome by \(o \in \{s,f\}\). The
probability of observing a success is increasing in \(\theta\) and in
\(\omega\). Whenever the agent observes a success, he gets a payoff
\(v>0\) and whenever the outcome is a failure, the payoff is 0. In
addition, the probability of success is such that for each state, there
is a unique optimal action that maximizes the agent's expected payoff.
Therefore, the probability of success can be seen as a monotone
transformation of the utility from Framework 1.

I focus on two nested theories that have been widely studied within this
framework: Full Bayesian updating and self-serving attribution bias. I
explain each of these classical models of belief updating in what
follows

\hypertarget{the-bayesian}{%
\subsubsection{The Bayesian}\label{the-bayesian}}

A Bayesian agent simultaneously updates their beliefs about \(\theta\)
and \(\omega\) by using Bayes' rule. The posterior at period t+1 about
\(\theta\) after observing outcom \(o\) is given by: \[
p_{t+1}(\theta, \omega| h^t) = \frac{p[o|\theta, \omega]p_t(\theta, \omega)}{\sum_{(\theta', \omega')}p[o|\theta', \omega']p_t(\theta', \omega')}
\]

Where \(p_{t}\) is the belief at the start of period \(t+1\) and
includes all the information gathered so far.

Bayesian agents choose the effort level that maximizes their expected
flow payoff by taking expectations over their prior beliefs about
\(\theta\) and \(\omega\). Since agents are myopic, even though all the
parameters could be identified with enough variation in choices, there
will be instances in which the Bayesian will not learn even with
infinite amounts of data. This happens because, by being myopic, they do
not internalize the tradeoff between flow payoff and learning. This can
result in too little experimentation to learn their true type. An
alternative to this approach is given by \citet{Hestermann2021} and is
discussed in the concluding remarks. Regardless, this approach is useful
as a benchamrk for the other theories discussed in this paper.

Also notice that if a fully Bayesian agent has a dogmatic prior, they
will never update their beliefs about the parameter that they are
dogmatic about. This will imply that they are prone to the same types of
errors as the dogmatic modeler in framework 1.

\hypertarget{the-self-serving-updater}{%
\subsubsection{The Self-Serving
Updater}\label{the-self-serving-updater}}

A Self-Serving Bayesian is an agent who uses a biased version of Bayes
rule to update their beliefs. They will update their beliefs about the
state \(\omega\) and his type \(\theta\) simultaneously by
over-attributing successes to a high value of \(\theta\) and
under-estimating the role of higher \(\omega\). Similarly, he will
attribute failure to a low state to a greater degree than an unbiased
agent would. To model the self-serving attribution bias, I take the
approach of \citet{benjamin2019}, where the posterior odds are given by:

\[
p_{t+1}(\theta, \omega| h^t) = \frac{p[h^t|\theta, \omega]^{c(\theta, \omega, o_t)}p_t(\theta, \omega)}{\sum_{(\theta', \omega')}p[h^t|\theta', \omega']^{c(\theta, \omega, o_t)}p_t(\theta', \omega')}
\]

with
\(c(\theta_H, \omega, o)<c(\theta_M, \omega, o)<c(\theta_L, \omega, o)\leq 1\)
and
\(c(\theta, \omega_L, o)<c(\theta, \omega_M, o)<c(\theta, \omega_H, o)\leq 1\).

This formulation of the bias means that the agent will over-attribute
successes to a high type and under-attribute them to a low type Whereas
they will over-attribute failures to a low state and under-attribute
them to a high state. The bias is introduced by distorting the perceived
likelihood that the signal was generated by either high values of
\(\theta\) when it is a success, or by low values of \(\omega\) when it
is a failure. The parameter \(c\) determines the degree of bias and
imposes enough restrictions to ensure that the model is still
falsifiable.

\hypertarget{a-unifying-example}{%
\section{A Unifying Example}\label{a-unifying-example}}

In order to compare the predictions of the theories discussed above, I
develop a unifying example that allows me to isolate the forces behind
each of the theories. The example is a modification of the one in
\citet{Heidhues2018} and is adapted to implementable in the lab.

The agent can be of one of 3 types:
\(\theta \in \{\theta_L, \theta_M, \theta_H\}\) with
\(\theta_H > \theta_M > \theta_L\). They face an unknown exogenous
success rate \(\omega \in \{\omega_L, \omega_M, \omega_H\}\) with
\(\omega_H>\omega_M>\omega_L\). Each of the values of \(\omega\) is
realized with equal probability. The agent knows the distribution of
\(\omega\) but not its realized value.

Denote the true type by \(\theta^*\) and the true state by \(\omega^*\).
The agent holds some prior belief about
\(\theta\)\footnote{which is potentially misspecified as in the dogmatic and switcher cases discussed above}
and chooses a binary gamble \(e in \{e_L, e_M, e_H\}\). The agent
observes whether the gamble is a success or a failure and gets a payoff
of \(1\) and if it is a success; they get \(0\) otherwise.

The probability of success is increasing in both \(\theta\) and
\(\omega\) and is fully described by the Table \ref{tab:DGP}

\begin{table}[htbp]
\centering

\begin{tabular}{ c|c|c|c|}

  \multicolumn{1}{c}{} & \multicolumn{1}{c}{$\omega_H$} & \multicolumn{1}{c}{$\omega_M$} & \multicolumn{1}{c}{$\omega_L$}\\
  \cline{2-4}
  $e_H$ & 50 & 20 & 2 \\
  \cline{2-4}
  $e_M$ & 45 & 30 & 7 \\
  \cline{2-4}
  $e_L$ & 40 & 25 & 20 \\

  \cline{2-4}
  \multicolumn{1}{c}{} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{$\theta_L$} & \multicolumn{1}{c}{}\\
\end{tabular}
\hspace{.3cm} 
\begin{tabular}{ c|c|c|c|}

  \multicolumn{1}{c}{} & \multicolumn{1}{c}{$\omega_H$} & \multicolumn{1}{c}{$\omega_M$} & \multicolumn{1}{c}{$\omega_L$}\\
  \cline{2-4}
  $e_H$ & 80 & 50 & 5 \\
  \cline{2-4}
  $e_M$ & 69 & 65 & 30 \\
  \cline{2-4}
  $e_L$ & 65 & 45 & 40 \\
  \cline{2-4}
  \multicolumn{1}{c}{} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{$\theta_M$} & \multicolumn{1}{c}{}\\
\end{tabular}
\hspace{.3cm} 
\begin{tabular}{ c|c|c|c|}

  \multicolumn{1}{c}{} & \multicolumn{1}{c}{$\omega_H$} & \multicolumn{1}{c}{$\omega_M$} & \multicolumn{1}{c}{$\omega_L$}\\
  \cline{2-4}
  $e_H$ & 98 & 65 & 25 \\
  \cline{2-4}
  $e_M$ & 80 & 69 & 35 \\
  \cline{2-4}
  $e_L$ & 75 & 55 & 45 \\
  \cline{2-4}
  \multicolumn{1}{c}{} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{$\theta_H$} & \multicolumn{1}{c}{}\\
\end{tabular}

\caption{Probability of success for each type, gamble and effort level}
\label{tab:DGP}
\end{table}

Conditional on a type, the agent's flow payoff is maximized by choosing
the gamble that matches the state. For example, if the value of
\(\omega\) is \(\omega_H\), the agent's flow payoff is maximized by
choosing \(e_H\) and if the state is \(\omega_L\) the flow payoff is
maximized by choosing gamble \(e_L\), regardless of the value of
\(\theta\). The agent myopically chooses gambles every period to
maximize the flow payoff for \(T<\infty\) periods.

After observing the outcome of each gamble, the agent updates their
beliefs using some procedure and moves on to the next period.

Notice that both \(\theta\) and \(\omega\) can be identified from the
outcomes if enough variation in the effort choices exists. This can be
seen by confirming that there is no pair of \(\theta\) and \(\omega\)
such that the probability of success is the same for all effort choices.
Thus, by changing the effort choice, the agent can learn both their type
and the state if they observe enough outcomes.

In this example, for an agent with a dogmatic belief about their type, a
self-defeating equilibrium is one in which the agent chooses an effort
level that, under the true \(\theta\), yields a frequency of success
that is consistent with the agent's misspecified belief. That is
\(P[\text{sucess}|\theta^*, e^*] = P[\text{sucess}|\hat{\theta}, e^*]\)
where \(e^*\) is the agent's myopic optimal choice.

In the data-generating process described above, there are five such
equilibria. For example, if the agent is of type \(\theta_M\) but
mistakenly believes that he is of type \(\hat{\theta}=\theta_H\) and the
and \(\omega^* = \omega_M\), when the effort chosen is \(e_L\), the
agent will observe a success with 45\% chance. Because the agent
dogmatically believes that their type is high, they will erroneously
conclude that the rate is \(\omega_L\). Under this belief, the optimal
action is \(e_L\) which will continue to generate successes with \(45%
\) probability, further reinforcing the incorrect belief. By doing so,
the agent forgoes the payoff from gamble \(e_M\) which would yield a
success with \(65\%\) chance.

By including self-confirming equilibria, the example captures the forces
from each of the updating mechanisms discussed in the previous section
and allows for the comparison of the main forces behind the theories.
For realizations of \((\theta, \omega)\) for which there are
self-confirming equilibria, the dogmatic agent will fall into the trap
whereas the switcher will be able to escape it. Similarly, an agent with
self-attribution bias will update their beliefs differently from an
unbiased Bayesian, leading them to choose different gambles. I exploit
such cases in order to test which model is a better fit for how subjects
behave in a laboratory
experiment.\footnote{because the setting does not match that of @Heidhues2018, there will be situations for which the theory 
does not provide a prediction. If such cases arise in the lab, they will not be used for the analysis. However, whether a misspecified 
belief persists or not for the switcher, depends highly on the realized history of signals that he gets.}
In what follows I explain the details of how this example was
implemented in the lab.

\hypertarget{experimental-design}{%
\section{Experimental Design}\label{experimental-design}}

I recruited 78 undergraduate subjects from the CESS lab at NYU who
participated in an in-person experiment. Sessions lasted approximately
45 minutes and subjects earned an average payment of \(\$22.6\). The
experiment was programmed using oTree \citep{otree}.

The experiment consisted of 2 treatments: the \emph{ego-relevant}
condition and the \emph{stereotype} condition. Subjects participated in
only one of the treatments. All subjects within a session participated
in the same treatment and the first 4 treatments were assigned the
ego-relevant condition; the rest were assigned to the stereotype
treatment. The tasks were identical across treatments, except for
parameter \(\theta\). In the ego-relevant condition \(\theta\) is the
subject's own performance in a quiz, while in the stereotype condition,
it is the performance of a randomly selected subject from another
session.

The experiment had 3 parts. In Part 1 subjects had 2 minutes to answer
as many multiple-choice questions as they could from a 20-question quiz.
They did this for quizzes on 6 different topics. The topics were: Math,
Verbal Reasoning, Pop-culture and Art, Science and Technology, US
Geography, and Sports and Video Games. In this part, they did not know
how many questions were available and they were given no feedback.

After taking all 6 quizzes, they proceeded to part 2 where they were
asked to guess their score on each of them. In the stereotype treatment
they were additionally asked to guess the score of a randomly drawn
participant from a previous session. All they knew about the other
participant was their gender identity and whether they were US nationals
or not. For each guess, they had three score options: Low-Score (5 or
fewer correct answers), Mid-Score (between 6 and 15 correct answers),
High-Score (16 or more). Each of the score categories corresponded to
\(\theta_L\), \(\theta_M\), and \(\theta_H\) respectively. They were
also asked to say how confident they felt about their choices. They had
4 possible answers: ``it was a random guess'', ``there is another
equally likely score'', ``I am pretty sure'', ``I am completely sure''.
These 4 answers are mapped to priors that place probabilities .33, .50,
.75, and 1 to the chosen type. The remaining probability is split
equally among the other two types. Questions in Part 2 were not
incentivized, but subjects were told that providing an accurate answer
would increase their chances of earning more money in the last part of
the experiment.

The purpose of Part 2 is to classify subjects into overconfident,
underconfident and correctly specified. If a subject guesses their score
to be in a higher (lower) category than their true score, they are
overconfident (underconfident); if they guess their score to be in the
same category as their true score, they are correctly specified. This
classification is done for each of the 6 topics separately.

Finally, in Part 3, subjects completed a belief updating task for each
of the quizzes. Before starting the task they were reminded of their
guess for the score. In the ego-relevant treatment, they were reminded
of their guess about themselves and in the stereotype treatment they
were reminded of their guess about the other participant. In the
stereotype treatment, they were also reminded of the characteristics of
the other participant.

For one topic at a time and in random order, they were presented with
the three gambles from the example above and were asked to choose one of
them. The probability of success was determined by their own score in
the ego-relevant condition, and by the score of the other participant in
the stereotype condition. Subjects had access to the three probability
tables in the printout of the instructions at all times and the meaning
of each cell was explained in detail.

In the interphase, they had to choose which of the 3 tables they wanted
to see before entering their choice in it. This was done as an
alternative to a belief elicitation in each round. I take their choice
of table to be indicative of their beliefs about the underlying type. I
chose not to elicit the beliefs at each round to stay true to the forces
in framework 1.

Once they have entered their choice, they observe a sample of 10
outcomes from the gamble they chose. After observing the outcomes, they
returned to the choice screen and entered a new choice. In the choice
screen subjects had access to the entire history of gambles and outcomes
for that task as well as a summary of the outcomes so far. Once they
entered 11 gambles (and observed 110 outcomes), they moved on to the
next topic and repeated the same procedure. They all did this for all 6
topics.

At the end of the experiment, one of the 6 topics was randomly selected
to determine the payment. They earned \(\$0.20\) for each correct answer
in the quiz, and for each success in the task in Part 3 for the selected
topic.

Randomness is controlled throughout the experiment and sessions by
setting a seed at the beginning of the first session. The seed was drawn
at random and remained fixed for all
sessions.\footnote{The seed that was drawn 
at the beginning of session 1 was 3452. The same seed was used for all sessions. It is used both for drawing $\omega$ for each of 
the tasks in the experiment, as well as for drawing the outcomes from the gambles.}
By doing this I ensure that any two subjects who have the same type and
face the same exogenous rate will observe the same outcomes. This
feature allows me to identify differences across subjects in updating
procedures since if they use the same updating procedure, they should be
choosing the same gambles.

\hypertarget{predictions}{%
\section{Predictions}\label{predictions}}

In this section, I outline the behavior that is predicted by each of the
theories discussed above.

\hypertarget{dogmatic-modeler}{%
\subsection{Dogmatic Modeler}\label{dogmatic-modeler}}

Since the domain of the problem is discrete and finite in the example,
the predictions of the original theory apply only to the combinations of
parameters and initial beliefs for which there is a self-defeating
equilibrium. In the example, there are 5 such combinations. For each of
them, the dogmatic modeler predicts that the agent will fall into the
self-defeating equilibrium and will be able to sustain the misspecified
belief forever.

Table \ref{tab:table-dog} describes the 5 self-defeating equilibria and
the effort choices that sustain them. The first columns describe the
combination of parameters and initial beliefs. The last column describes
the effort choice that the agent will make in the long run. It is only
for these combinations of parameter values and beliefs that the dogmatic
model makes predictions within the Unifying example.

\textbf{Prediction 1D:} \emph{Whenever an agent is of a type
\(\theta^*\) but mistakenly believes that they are of a type
\(\hat{\theta}\), and \((\theta^*, \omega^*, \hat{\theta})\) are such
that there is s self-defeating equilibrium, the agent will fall into the
trap and choose the effort level that sustains the misspecified belief
forever.}

The model does not make predictions about what happens in cases where
there is no stable belief. I assume that because there is no stable
belief, the agent will eventually have to use some procedure to revise
their belief about \(\theta\). In such cases, I aim to determine which
of the alternative explanations provided by the other theories is a
better fit for the data.

\begin{table}[htbp]

    \centering
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \textbf{True Type ($\theta^*$)} & \textbf{True State ($\omega^*$)} & \textbf{Believed Type ($\hat{\theta}$)} & \textbf{Believed state ($\hat{\omega}$)} & \textbf{Effort} \\
        \hline
        \hline
        $\theta_L$ & $\omega_H$ & $\theta_M$ & $\omega_L$ & $e_L$ \\
        \hline
        $\theta_M$ & $\omega_L$ & $\theta_L$ & $\omega_M$ & $e_M$ \\
        \hline
        $\theta_M$ & $\omega_M$ & $\theta_H$ & $\omega_L$ & $e_L$ \\
        \hline
        $\theta_M$ & $\omega_M$ & $\theta_L$ & $\omega_H$ & $e_H$ \\
        \hline
        $\theta_M$ & $\omega_H$ & $\theta_H$ & $\omega_M$ & $e_M$ \\
        \hline

    \end{tabular}
    \caption{Stable beliefs and the effort choices that support them for the unifying example}
    \label{tab:table-dog}
\end{table}

Although the dogmatic model does not apply to all possible
parametrizations and beliefs, whether subjects fall into the
self-defeating equilibria or not is still informative of the updating
procedure that they are using. Understanding if the presence of traps is
a key feature preventing subjects from learning the optimal action is
important for understanding the prevalence of overconfidence. Similarly,
gaining insight into what happens when there are no traps is important
for understanding what are the other reasons why overconfidence might
arise and prevail.

\hypertarget{switcher}{%
\subsection{Switcher}\label{switcher}}

Since the switcher starts as the same dogmatic agent, the initial
behavior of both types of agents is identical, however, because the
switcher is keeping track of the likelihood ratio, they will be able to
escape the self-defeating equilibrium if the evidence is convincing
enough. Therefore there is a positive probability that the switcher will
adjust their initially misspecified belief about \(\theta\) and learn
the true state.

\textbf{Prediction 1S:} With positive probability, the switcher will
escape the self-defeating equilibrium and learn the true state.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=2.5cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

@@

\newcommand{\EE}[1]{\mathbb{E}\left[#1\right]}
\newcommand{\PP}[1]{\mathbb{P}\left(#1\right)}
\newcommand{\CP}[2]{\mathbb{P}\left(#1 \,| \, #2 \right)}
\newcommand{\CE}[2]{\mathbb{E}\left[#1\,|\,#2 \right]}
\usepackage[bottom]{footmisc}
\usepackage[doublespacing]{setspace}
\usepackage[normal]{caption}
\usepackage{dsfont}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{hyperref}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Learning with Misspecified Models: the case of overestimation},
  pdfauthor={Jimena Galindo},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Learning with Misspecified Models: the case of overestimation}
\author{Jimena Galindo}
\date{October 15, 2023}

@@

  \bibliography{references.bib}
